---
title: "STAT 545B Assignment B4 - Andrea Wong Koo: Option A"
output: github_document
date: "2024-12-5"
---

``` {r}
library(tidyverse)
library(stringr)
library(janeaustenr)
library(ggplot2)
library(tidytext)
library(RColorBrewer)
library(datasets)
library(forcats)
library(OVtool) #for `sud` dataset
library(broom)
library(cowplot)

```


# Exercise 1: 

_Take a Jane Austen book contained in the `janeaustenr` package, or another book from some other source, such as one of the many freely available books from [Project Gutenberg](https://dev.gutenberg.org/) (be sure to indicate where you got the book from). Make a plot of the most common words in the book, removing “stop words” of your choosing (words like “the”, “a”, etc.) or stopwords from a pre-defined source, like the `stopwords` package or `tidytext::stop_words`._

_If you use any resources for helping you remove stopwords, or some other resource besides the `janeaustenr` R package for accessing your book, please indicate the source. We aren’t requiring any formal citation styles, just make sure you name the source and link to it._

- Book source: https://cran.r-project.org/web/packages/janeaustenr/janeaustenr.pdf 

``` {r}
# To create an object for one of the books in the `janeaustenr` package:
book_text <- austen_books() %>%
  filter(book == "Pride & Prejudice") %>% 
  select(text)

head(book_text)

```
``` {r}
# To explore the data structure of the stop words available in tidytext: 
glimpse(tidytext::stop_words)

```

- `tidytext::unnest_tokens()` source: https://www.rdocumentation.org/packages/tidytext/versions/0.4.2/topics/unnest_tokens 

Using `tidytext` and `purrr` functions, the code below 
``` {r}
# To produce and count frequencies of words in book, with stop words removed: 
common_words <- book_text %>%
  
  # To tokenize the text into individual words: 
  tidytext::unnest_tokens(output = book_word, input = text) %>% 
  
  # Remove stop words by filtering out words that appear in the 'stop_words' dataset:
  filter(!purrr::map_lgl(book_word, ~.x %in% stop_words[["word"]])) %>%
  
  # Count the occurrences of each unique word, sorted in descending order: 
  count(book_word, name = "count", sort = TRUE)

print(common_words)

``` 
To gain an idea of the gain idea of the distribution of word frequencies with summary statistics: 

``` {r}

# To produce summary statistics:
common_words_summary <- common_words %>%
  summarize(
    mean_count = mean(count),
    max_count = max(count),
    min_count = min(count),
    median_count = median(count),
    q1 = quantile(count, 0.25),
    q3 = quantile(count, 0.75),
    .groups = "drop"  # In case of grouping, ensures it doesn't propagate
  )

print(common_words_summary)
```
``` {r}
# To plot the distribution plot for word counts:
word_distribution <- ggplot(common_words, aes(x = count)) +
  geom_histogram(bins = 30, fill = "skyblue", alpha = 0.7) +
  labs(
    title = "Distribution of Word Frequencies",
    x = "Word Frequency Count",
    y = "Frequency",
    caption = paste("Mean: ", round(common_words_summary$mean_count, 2), 
                    " | Median: ", round(common_words_summary$median_count, 2))
  ) +
  theme_minimal()

print(word_distribution)

```


Based on the distribution of word frequencies, it seems like it is right skewed, with very high counts, it would be acceptable to choose the top 10 most frequently used words. 

``` {r}
# To view the top 10 most common words with the highest counts:
top_10_common_words <- head(common_words, 10) # Returns the top 10 rows

print(top_10_common_words)

```

Below is a visualization of the counts for the top 10 most common words. Based on the graph, the most common word is 'elizabeth' with a 600 word frequency count. 

``` {r}
# To visualize the counts of the top 10 most common words: 
common_words_plot <- top_10_common_words %>%
  ggplot(aes(x = reorder(book_word, count), y = count, fill = book_word)) +
  geom_bar(stat = "identity") + # Create bar plot for each common word
  coord_flip() + # For readability, flip coordinate to horizontal orientation: 
  labs(x = "Word", y = "Count", title = "Top 10 Most Common Words") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3") + # Uses a colour palette for bars
  theme(legend.position = "none") # Removes the legend
 
print(common_words_plot)
  

```


# Exercise 3

_For this exercise, you’ll be evaluating a model that’s fit separately for each group in some dataset. You should fit these models with some question in mind._

This data analysis uses a synthetic data set called `sud` from library(Ovtool), derived from a large-scale observational study on 4000 youth in substance use disorder treatment. The specific names of two different types of treatments are not specified, but are named "Treatment A" and "Treatment B". 

For this data analysis, we are interested in abstinence duration recorded at six months for treatment A, which is a treatment outcome. To analyze and predict this outcome of interest, we will specifically look at the number of treatment A procedures delivered to a client.

**Dataset source:** https://cran.r-project.org/web/packages/OVtool/vignettes/OVtool.html 

**Objective:** To build a linear regression model of number of abstinent days at a 6-month check point, based on number of treatment procedures for each substance use subgroup.

**Research Question:** How does the number of treatment A procedures affect and predict abstinence within 6 months of treatment within different substance use groups?

To carry out this data analysis, follow along with the steps.


## Question 1 

_Question 1: Make a column of model objects. Do this using the appropriate mapping function from the purrr package. Note: it’s possible you’ll have to make use of nesting, here._

First, we explore the documentation and data structure of the `sud` dataset to understand the variables available, data types of each variable, and a preview of the values.

``` {r}
# To view documentation of dataset:
?sud
```


``` {r}
# To preview the dataset values:
head(sud)

```

``` {r}
# To view the data structure of dataset including variables and data types: 
glimpse(sud)
```


After referring to the dataset source link and reviewing the data structure, the relevant variables for this analysis are: 

- **Outcome of interest:** `ada_6`: adjusted days abstinent (ADA) recorded at 6-months, a continuous numeric variable that functions as a treatment outcome. 

- **Predictor variable/variable of interest:** `npoc`: the count of Treatment A procedures delivered to client, a continuous numeric variable. 

- **subsgrps_n**: Since the dataset includes individuals who use different drugs, it would make sense to group the linear regression by substance subgroups to compare the effect of treatments on abstinence specific to each substance. Hence, the data analysis will group by `subsgrps_n`, which is a categorical variable where 1="Alcohol and/or marijuana disorder/weekly use; 2="Other drugs"; 3="Opiate disorder/weekly use".

Now that we have explored the variables in the dataset, we will make a column of model objects, where each row is a linear regression model, with Y outcome variable as `ada_6` and X predictor variable as `nproc`. 

``` {r}

# To create a column of linear model objects: 
sud_models <- sud %>%
  select(subsgrps_n, nproc, ada_6) %>% # To select relevant variables: 
  drop_na() %>%  # Removes missing values for linear regression functionality
  mutate(as_factor(subsgrps_n)) %>% # Ensures that `subsgrps_n` values are factors
  group_by(subsgrps_n) %>% # Group by subsgrps_n to fit models separately for each group
  nest(data = c(nproc, ada_6)) %>% # Nest data for each subsgrps_n groups
  mutate(model = map(data, ~ lm(ada_6 ~ nproc, data = .))) # Fitting the linear model

head(sud_models)

```

## Question 2: 

_Question 2: Evaluate the model in a way that interests you. But, you should evaluate something other than a single number for each group. Hint: you’ll need to use another purrr mapping function again._

To both analyze trends and predict abstinence given a number of treatment A procedures, we will evaluate both the regression coefficients using `broom::tidy`, and fitting values with `broom::augment`.

``` {r}
# To evaluate the regression coefficients (tidy output): 
sud_lm_tidy <- sud_models %>%
  mutate(tidy_output = map(model, broom::tidy))

```


``` {r}
# To fit values and residuals (augment output): 
sud_lm_augment <- sud_models %>%
  mutate(augmented_data = map(model, broom::augment, .keep = "none"))

```

## Question 3: 

_Question 3: Print out this intermediate tibble for inspection (perhaps others as well, if it makes sense to do so)._

Below are the outputs of the model evaluations:

``` {r}
# Prints the tidy output: 
print(sud_lm_tidy)
```

``` {r}
# Prints the augment output: 
print(sud_lm_augment)
```

## Question 4: 

_Question 4: Unnest the resulting calculations, and print your final tibble to screen. Make sure your tibble makes sense: column names are appropriate, and you’ve gotten rid of columns that no longer make sense._

To view the results of the models, we can unnest and only include the most informative columns.

### Tidy output: 

**Tidy output interpretation:** For the tidy output,  The `term` column includes the intercept and the coefficient for nproc, with the corresponding `estimate`, `std.error`, `statistic`, and `p.value` columns.  The `intercept` represents the starting point of the regression line. The `estimate` represents the effect of the predictor on the outcome, the `std.error` is the standard error of the estimate, statistic is the test statistic (e.g., t-value), and `p.value` indicates whether the relationship is statistically significant.

Comparing the linear regression coefficients, the Alcohol and/or marijuana disorder/weekly use group (group 1) shows a statistically significant positive relationship between the number of treatment procedures (`nproc`) and the number of days abstinent at 6-month checkpoint (`apa_6`). The coefficient of 0.13 with p < 0.001, suggests that more treatment procedures correlate with a slightly higher likelihood of abstinence.

In contrast, the Other drugs (group 2)and Opiate disorder/weekly use group (group 3) do not show a statistically significant relationship between the number of treatment procedures and abstinence, with very small and non-significant coefficients of 0.01 and 0.06, respectively with p > 0.5.


``` {r}

# To unnest the model tidy output:
final_sud_tidy <- sud_lm_tidy %>%
  unnest(tidy_output) %>%  # Unnest the list-column containing the tidy output
  select(subsgrps_n, term, estimate, std.error, statistic, p.value)  # Keep relevant columns

print(final_sud_tidy)

```

### Augment model: 

To view the fitted values and residuals of predictions in the linear regression, we can unnest the model and only include the relevant columns, as shown below:

``` {r}
# To unnest the model augment output: 
final_sud_augment <- sud_lm_augment %>%
  unnest(augmented_data) %>% # Unnest the list column
  select(subsgrps_n, ada_6, nproc, .fitted, .resid) %>% # Only keep columns of interest
  rename(  # Renamed variables for interpretability
    fitted_values = .fitted,
    residuals = .resid)

print(final_sud_augment)
```

## Question 5: 

_Question 5: Produce a plot communicating something about the result._

### Tidy output plot: 

**Plot interpretation:** Below is a regression coefficient plot, which allows you to visually assess the strength and precision of the relationship between `nproc` (number of treatment A procedures) and the outcome variable `ada_6` (number of days abstinent at 6 months) within each substance use subgroup. The error bars identify the level of uncertainty in each estimate. Subgroups with narrow error bars and estimates far from zero are likely to show a stronger and more reliable relationship. Conversely, subgroups with wide error bars and estimates near zero indicate a weaker or less certain relationship.

In this plot, the points for `nproc` for each substance use group is shown, located on their corresponding estimates with error bars. 

**Conclusion:** Based on the regression coefficients from the `broom::tidy output`, the results suggest that treatment A procedures may have a more positive effect on abstinence in the Alcohol and/or marijuana disorder group, as indicated by a positive regression coefficient. However, the long error bar (i.e., large standard error) for this group suggests significant uncertainty in the estimate, meaning that the effect is not certain. Additionally, the minimal impact observed in the other groups (Other drugs and Opiate disorder/weekly use) could be due to either a smaller effect or greater uncertainty, as reflected by the standard errors and p-values, indicating that the conclusions for these groups should be taken with caution.


``` {r}
# Filter for 'nproc' term only:
final_sud_tidy_nproc <- final_sud_tidy %>%
  filter(term == "nproc")  # Only include rows where term is 'nproc'

# Adds a custom label for each group
final_sud_tidy_nproc <- final_sud_tidy_nproc %>%
  mutate(term = paste0("nproc_", subsgrps_n))  # Modify term names to reflect group names


# To specify labels for the legend
group_labels <- c(
  "1" = "Alcohol and/or marijuana disorder", 
  "2" = "Other drugs", 
  "3" = "Opiate disorder"
)

# Regression coefficient plot from linear regression model (from tidy) showing only 'nproc': 
sud_tidy_plot <- final_sud_tidy_nproc %>% 
  ggplot(aes(x = estimate, y = term, color = factor(subsgrps_n))) +  # Colour by group
  geom_point(alpha = 1) + 
  geom_errorbarh(aes(xmin = estimate - std.error, xmax = estimate + std.error), height = 0.2, alpha = 1) + # Horizontal error bars based on standard error
  labs(
    title = "Regression Coefficients for nproc (Number of Treatment Procedures)",
    x = "Estimate with Standard Error",
    y = "nproc for each Substance Use Subgroup",
    color = "Substance Use Group"  # Title for the colour legend
  ) +
  scale_color_manual(values =c("1" = "#6A8DFF", "2" = "#FF6B6B", "3" = "#8FD9A8"),
                     labels = group_labels) +  # Custom legend labels
  theme_minimal() +
  guides(color = guide_legend(title = "Substance Use Group (weekly use)"))  

print(sud_tidy_plot)

```

### Augment output plot: 

**Plot interpretation:** The scatter plot illustrates the relationship between the number of Treatment A procedures (`nproc`) on the x-axis and the adjusted number of days abstinent at 6 months (`ada_6`) on the y-axis. Each point represents an individual observation, color-coded by substance use subgroup. The linear regression lines depict the predicted relationship between these variables, adjusting for each subgroup independently. The fitted lines provide insights into how the relationship between treatment procedures and abstinence varies across different substance use groups.

From the plot, we can observe that all three groups show an upward trend in the regression lines, indicating that an increase in the number of Treatment A procedures is generally associated with more abstinent days at the 6-month mark. However, the steepness of the slopes differs. The Alcohol and/or Marijuana Use group shows the steepest slope, suggesting a relatively stronger correlation, while the Other Drugs group has the flattest slope, indicating a weaker relationship. The Opiate Disorder group lies in between, reflecting a moderate correlation. The wide spread of data points across all groups highlights considerable variability in how the number of procedures correlates with abstinence.

**Conclusion:** The linear regression plot indicates a positive correlation between the number of Treatment A procedures and the number of days abstinent, with variations in strength across the substance use groups. The Alcohol and/or Marijuana Use group shows the strongest association, as evidenced by the steepest regression slope. However, the larger density of data points in this group may suggest that the observed correlation is partly influenced by the greater number of individuals in this subgroup. Despite the upward trend, the wide distribution of data points in all groups suggests that the linear relationship between treatment procedures and abstinence is not perfectly consistent. This variability calls for caution in interpreting the strength of the correlation, as individual factors may play a significant role.

``` {r}
# Scatter plot with fitted values and linear regression model (from augment):
sud_augment_plot <- final_sud_augment %>%
  ggplot(aes(x = nproc, y = ada_6, color = as_factor(subsgrps_n))) +
  geom_point(alpha = 0.6) + # Scatter plot of points 
  geom_line(aes(y = fitted_values), linewidth = 0.5) + # Linear regression line 
  facet_wrap(~ subsgrps_n, scales = "free") + # Group plots by substance groups 
  labs(
    title = "Adjusted Days Abstinent at 6 months vs. Treatment A Procedures Delivered", 
    x = "Number of Treatment A Procedures",
    y = "Adjusted Days Abstinent (ADA) at 6-months",
    color = "Substance Subgroup"
  ) +
  scale_color_manual( # To add colours and legend labels for each group 
    values = c("1" = "#6A8DFF", "2" = "#FF6B6B", "3" = "#8FD9A8"),
    labels = c(
      "1" = "Alcohol and/or marijuana disorder",
      "2" = "Other drugs",
      "3" = "Opiate disorder"
    )  # Specify labels for the legend
  ) +
  theme_minimal() +
  theme(legend.position = "bottom") # To place the legend at the bottom 

print(sud_augment_plot)


```

### Overall conclusion:

In this analysis, we explored the relationship between the number of Treatment A procedures and the number of abstinent days at the 6-month checkpoint in youth undergoing substance use disorder treatment. Linear regression models were fitted for each substance use subgroup: Alcohol and/or marijuana disorder, Other drugs, and Opiate disorder. The results indicate that for the Alcohol and/or marijuana disorder group, a statistically significant positive relationship was observed, with each additional treatment procedure slightly increasing abstinence duration. However, for the Other drugs and Opiate disorder groups, no significant relationship was found, suggesting that the number of treatment procedures may not strongly predict abstinence in these subgroups. This is further supported by the regression coefficients and p-values, where large standard errors and non-significant p-values highlight uncertainty or minimal effect in these groups. Thus, while Treatment A may have a positive impact on abstinence in some subgroups, further investigation is needed to understand the variability in response across different substance use disorders, perhaps by conducting analysis or regression with more variables. 



